Assessing the reliability of an online behavioural laterality battery: a pre-registered study.

### 2020-08-07 online_laterality_reliability.Rmd
used for pre-print

---
title: 'Assessing reliability of an online behavioural laterality battery: a pre-registered
  study.'
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

<!--- updated 20/7/20 to change 'select' to 'dplyr::select' to avoid conflicts-->
<!--- updated 3/8/20 to add ZW feedback--->
Adam J. Parker

@@

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. Online testing can potentially allow for much larger sample sizes, but limits the amount of experimental control that is feasible. Here we considered whether online testing could give valid and reliable results on four tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task to assess manual preference. We recruited 392 participants, oversampling left-handers, who completed the battery twice. Three of the tasks showed evidence of both validity and reliability, insofar as they showed hemispheric advantages in the expected direction and test-retest reliability of at least *r*= .65. The reliability of the rhyme judgement task was less satisfactory (*r*= .63). We also found that left-handers were more likely to depart from typical lateralisation. Contrary to prediction, lateralisation across tasks was not significantly correlated. We conclude that three of the tasks, dichotic listening, chimeric faces and finger tapping, show considerable promise for online evaluation of cerebral lateralisation, but the results do not support the notion of a common underlying cause of lateralisation for the different cognitive tasks studied here.

@@

\newpage
## Introduction

Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well-powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. With respect to the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). Reliability can either be assessed within a test session (e.g. split-half reliability, from correlations between odd and even items) or from repeated assessments on different occasions (test-retest reliability). The latter is generally preferred as it includes all test items and will incorporate fluctuations in performance due to changes in state or test conditions from one occasion to another. Assessment of test-retest reliability, however, requires large samples of people who undergo repeated testing, and this is seldom feasible for in-person testing. Online testing allows for recruitment of much larger samples, but raises questions about test validity when opportunities for experimental control are reduced. Here we introduce an online behavioural laterality battery for which we established the test-retest reliability using a sample of 392 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. Validity of the tasks was assessed in terms of whether lateralisation was observed in the predicted direction, based on prior literature with in-person testing on analogous tasks. We also examined two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right handers on each behavioural laterality task? (2) do the laterality indices from different tasks correlate?

@@

In this paradigm, visual stimuli are briefly presented to the left or right of a central fixation stimulus. The visual pathways from the left and right sides of space to visual cortex are crossed for non-foveal viewing, and so the right visual-field (RVF) projects directly to the left hemisphere, and the left visual-field (LVF) to the right hemisphere. The corpus callosum allows for transfer between hemispheres, but the rationale of the method is that there will be a processing advantage for language stimuli that are directly projected to the language-dominant hemisphere. This method has been used since the 1950s (Mishkin & Forgays, 1952). Bourne (2006) reviewed the literature, which consistently shows advantageous processing for letter strings presented in the RVF, relative to the LVF, in most people.

Despite wide reporting of a RVF advantage at the population level,  the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field task consisted of lexical decisions made towards abstract nouns presented in the LVF and RVF. During the experiment, the blood flow to left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task— the gold standard task for assessing hemispheric dominance for language. Despite showing the expected RVF advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (*r*= .18) or the visual half-field task itself (r -.004). The lack of validity here likely reflects the poor reliability of the visual field task as split-half reliability was very low (reaction time spilt-half: *r*= .14; accuracy split-half: *r*= .20). This illustrates the difficulty of drawing conclusions about associations between laterality indices when reliability of measurement is so low.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Brysbaert and Hunter argued that this is the result of stimuli competing during bilateral presentation, enhancing the advantage to the stimulus presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation to prevent eye movements. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) used two visual half-field tasks, word naming (192 trials) and picture naming (160 trials) in which stimuli were bilaterally presented and masked at offset. The two tasks correlated strongly (*r*= .8, 95% CI .598-.956) in a sample of 26 left-handers. Encouragingly, the LIs from these visual half-field tasks also correlated positively with laterality from functional Magnetic Resonance Imaging (fMRI) on a word generation task in 10 left-handers. However, the small sample size meant that the confidence intervals around these estimates were very wide: word naming, *r*= .63 (95% CI = .001-.902), picture naming, *r*= .77 (95% CI = .272-.943).

@@

Simultaneous presentation of different auditory stimuli to left and right ears for immediate recall was a method originally used in studies of attention, but it became evident that there was usually an advantage for verbal stimuli presented to the right ear (Kimura, 1967). Both ears project to both hemispheres, but the weaker ipsilateral pathways appear to be suppressed when the two sides are in competition, and so the right ear, with a contralateral projection to the left hemisphere, will have an advantage in tasks that involve language processing.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. Their recommendations were to use: (1) stop consonant-vowel (CV) syllables as stimulus materials; (2) CV stimulus pairs from the same voicing category; (3) alternating trials of voiced and unvoiced stimulus pairs; (4) free-recall instruction; (5) onw stimulus pair per trial; (6) a paradigm length of 120 dichotic trials; (7) all stimulus pairs are presented to the two sides in identical frequency; and (8) to include binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded a test-retest reliability of *r*= .93 for verbal responses and *r*= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

@@

In studies of face processing, it is possible to use a modification of the visual half-field paradigm in which single stimuli are viewed, composed of vertically split chimeric faces, where each half displays a different emotion or identity. Our focus here is on emotion detection in chimeric faces, where a bias is usually found towards reporting the emotion shown in the left hemiface.

The reliability of the chimeric faces tasks seem promising. Early work by Levy et al (1983) using two vertically presented chimeric faces, found split-half reliability of *r*= .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data, but in a task of recognising facial identity rather than emotion. In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognize centrally presented chimeric faces, which presented different identities in the right and left visual fields. It is encouraging to note that the behavioural measure significantly correlated with the fMRI laterality index (LI) calculated from on activation in the fusiform face area (*r*= .49). However, the sample size used here was small: with 17 participants, the 95% confidence interval around the observed correlation of .49 ranged from .012 to .786.

@@

The studies reviewed so far have indicated consistent population biases in lateralisation on behavioural tasks, but also there are also substantial individual differences. One factor that has been associated with the degree of lateralisation is handedness (for a comprehensive review, see Carey & Johnstone, 2014). Neuroimaging data indicates that for language tasks these differences in lateralisation primarily reflect the contribution of a subset of strong left-handers who are right hemisphere dominant for language, rather than indicating that this relationship operates on a continuum (Mazoyer et al., 2014). This association can be used to validate behavioural measures of lateralisation: we would expect to see departure from typical lateralisation in strong left-handers.

With regards to visual half-field tasks for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Weekes, Capetillo-Cunliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well powered study involving 100 participants, conducted by Willemin et al. (2016), reported no hint of any effect of handedness, when coded as left and right, or when divided into left (N = 19), mixed (N = 24) and right-handed (N = 57) groups, on the visual field advantage during a lexical decision task. This result illustrates the recurring challenge of interpreting null results in this field: do they reflect a true lack of relationship, poor test reliability, inadequate power to detect small effects, or specific methodological factors? We may note that Willemin et al. asked participants to respond via manual response, whereas other studies have involved an active speech production component, which could influence task lateralisation.

@@

For chimeric faces tasks, which typically show a left visual field advantage, there again appears to be a greater departure from the typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Levy et al., 1983; Rozkowski & Snelbecker, 1982). This was also studied by Karlsson et al. (2019). When viewing a single face with a different emotion shown in either hemiface, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task that involved two chimeric faces arranged vertically, a similar pattern emerged. 73% of right-handers and 57% of non-right-handers showing a leftwards bias.

From the above reviewed studies, it appears that handedness is related to laterality across a range of tasks (for further discussion see Carey & Johnstone, 2014). This relationship points towards a general laterality factor which influences lateralisation across tasks and behaviours, as predicted by several theoretical accounts (Behrmann & Plaut, 2015; Cook, 1984; Hellige, 1983; Kosslyn, 1987). In a study of left-handers using fMRI with five tasks, Gerrits, Verhelst, and Vingerhoets (2020) found that the just under half the participants showed either standard laterality (left-lateralised for language and praxis, right-lateralised for face memory, line-bisection and vocal emotion) or a complete reversal of this pattern, providing clear evidence against the idea that these functions are lateralised independently. It is also clear that association is far from perfect,  (Badzakova-Trajkov, Haberling, Roberts, & Corballis, 2010) and some studies have found no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). Our ultimate goal is to unpick whether dissociations between laterality indices are evidence of multiple independent lateral biases across tasks, or whether there is a more prosaic explanation in terms of poor reliability or validity of tasks.

@@

The principal goal of the current study was to evaluate the reliability and validity of an online behavioural laterality battery. Specifically, we examined whether it is possible to observe the usual laterality biases at the population level in tests of language and emotion processing, and whether it is possible to achieve a test-retest reliability of *r*= .65 or above using online behaviour research methods. In addition, for tasks where satisfactory reliability was achieved, we considered how many trials were needed to achieve this. The tasks included in the battery were a novel rhyme decision visual half-field task, a consonant-vowel dichotic listening task, a chimeric faces visual half-field task, and a novel finger tapping task <!--- ZW: Would it be worth having a section on measurement of manual preference?; AP: As this is exploratory I don't want to add too much about this--->. These behavioural tasks were completed twice on two different occasions.

Our first set of predictions (hypothesis 1) concerned validity of the tasks as indicators of cerebral lateralisation. Based on previous work, we predicted that at the population level: (1) the laterality indices for rhyme decision would show a RVF advantage (indicating left hemisphere dominance for language processing); (2) the laterality indices for the dichotic listening task would show a right ear advantage (indicating left hemisphere dominance for language prcessing); and (3) the laterality indices for chimeric faces would show a LVF advantage (indicating right hemisphere dominance for face or emotion processing). Although exploratory, the finger tapping task was designed to provide a quantitive measure of degree of handedness, and was expected to show a right hand advantage in the overall sample.

Our second prediction (hypothesis 2) related to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotients of -90 or less on the Edinburgh Handedness Inventory; Oldfield, 1971) would be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population level predictions) than right-handers. This prediction followed from observations by Mazoyer et al (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded “concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

Our final prediction (hypothesis 3) related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of the individual measures. In the preregistration for our current study, we predicted that for right-handers, a model where the two language tasks are correlated with each other but not with the chimeric faces (model A) would be the best fit to the data, compared to either a model in which laterality indices are correlated between the language and chimeric face tasks (model B), or a base model where each task is independently lateralised (model C). Given our previous findings using fTCD to study dissociations in language laterality (Woodhead, Bradshaw, Wilson, Thompson, & Bishop, 2019), we further predicted that model fit may differ for right- and left-handers, with more dissociation between tasks in the latter group.

@@

This project was pre-registered on the Open Science Framework (OSF) prior to the start of data collection. The registration form along with the analysis scripts <!-- And task stimuli? Or will tasks be available on Gorilla?---> and anonymised data can be found on the OSF at https://osf.io/qsxcv/.

@@

In practice, our final sample exceeded 300, because we discarded participants from individual tasks for failing to meet prespecified criteria (see below). In total, 441 native English speakers were initially recruited through Prolific Academic (www.prolific.co). Participants had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss orimpairment.

This study was approved by the University of Oxford's Medical Sciences Inter-Divisional Research Ethics Committee (reference number: R66638/RE002) and conducted in accordance with the principles of the Declaration of Helsinki. Participants gave informed consent and received compensation for their participation.

@@

The battery included four measures of background information and four experimental assessments of behaviural laterality that were built and administered using the Gorilla Experiment Builder (www.gorilla.sc), a cloud-based tool for collecting data in the behavioural sciences (Anwyl-Irvine, Massonnié, Flitton, Kirkham, & Evershed, 2020). Each task is described below.

@@

The basic demographics questionnaire asked participants to report age, gender, years in education, and whether they were bilingual. We also asked questions about handedness and footedness. For handedness we asked: which is your dominant hand (i.e. which hand do you prefer to use for tasks such as writing, cutting, and catching a ball)? For footedness we asked: which foot do you normally use to step up on a ladder/step? For both questions, participants could answer "left", "right", or "no preference".

@@

The Edinburgh Handedness Inventory (EHI; Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The EHI was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference. For simple contrast between left- and right-handers, we treated indices greater than 0 as right handed, and indices less than 0 as left handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.

@@

The rhyme decision task was developed for this project as a novel visual half-field technique designed to engage lateralised language processes. On each trial, participants are asked to decide whether a foveally presented letter string rhymed with the name of an image shown in either the LVF or RVF. With online presentation, a particular concern is potential for eye movements away from fixation. We relied on the fact that participants had to focus initially on the central written word in order to do the task, with pictures subsequently displayed with brief bilateral presentation, which has been shown to give adequate control for eye movements in previous visual half-field experiments with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements in this setting reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972; but see also Bourne, 2006). Pictures of familiar objects were presented for rhyme matching to the written word, with the goal of ensuring that the participant would have to covertly name the picture, thereby engaging left hemisphere speech processing.

@@

![Figure 2. An example stimulus from the chimeric faces task. In the example, “happiness” is shown on the left and “sadness” on the right. A participant who is right lateralised for emotional processing, would be more likely to perceive the emotion in the left hemifield and would therefore respond “happiness”.](chim_example.png)

@@

autofit(ft)
```
*Note*. N is the number of participants. % is the column percentage of participants.

@@

#### Supplemental material

>>>>

---
title: 'Assessing the reliability of an online behavioural laterality battery: a pre-registered study.'
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

<!--- updated 20/7/20 to change 'select' to 'dplyr::select' to avoid conflicts-->
<!--- updated 3/8/20 to add ZW feedback--->
Adam J. Parker*

@@

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. Online testing can allow for much larger sample sizes but limits the amount of experimental control that is feasible. Here we considered whether online testing could give valid and reliable results on four tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task. We recruited 392 participants, oversampling left-handers, who completed the battery twice. Three of the tasks showed evidence of both validity and reliability, insofar as they showed hemispheric advantages in the expected direction and test-retest reliability of at least *r*= .75. The reliability of the rhyme decision task was less satisfactory (*r*= .63). We also found that left-handers were more likely to depart from typical lateralisation. Contrary to prediction, lateralisation across tasks was not significantly correlated. We conclude that three of the tasks, dichotic listening, chimeric faces and finger tapping, show considerable promise for online evaluation of cerebral lateralisation, but the results do not support the notion of a common underlying cause of lateralisation for the different cognitive tasks studied here.

@@

\newpage
Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well-powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. Concerning the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). Reliability can either be assessed within a test session (e.g. split-half reliability, from correlations between odd and even items) or from repeated assessments on different occasions (test-retest reliability). The latter is generally preferred as it includes all test items and will incorporate fluctuations in performance due to changes in state or test conditions from one occasion to another. Assessment of test-retest reliability, however, requires large samples of people who undergo repeated testing, and this is seldom feasible for in-person testing. Online testing allows for the recruitment of much larger samples but raises questions about test validity when opportunities for experimental control are reduced. Here we introduce an online behavioural laterality battery for which we established the test-retest reliability using a sample of 392 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. the validity of the tasks was assessed in terms of whether lateralisation was observed in the predicted direction, based on prior literature with in-person testing on analogous tasks. We also examined two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right-handers on each behavioural laterality task? (2) do the laterality indices from different tasks correlate?

@@

In this paradigm, visual stimuli are briefly presented to the left or right of a central fixation stimulus. The visual pathways from the left and right sides of space to the visual cortex are crossed for non-foveal viewing, and so the right visual-field (RVF) projects directly to the left hemisphere and the left visual-field (LVF) to the right hemisphere. The corpus callosum allows for transfer between hemispheres, but the rationale of the method is that there will be a processing advantage for language stimuli that are directly projected to the language-dominant hemisphere. This method has been used since the 1950s (Mishkin & Forgays, 1952). Bourne (2006) reviewed the literature, which consistently shows advantageous processing for letter strings presented in the RVF, relative to the LVF, in most people.

Despite wide reporting of a RVF advantage at the population level, the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field task consisted of lexical decisions made towards abstract nouns presented in the LVF and RVF. During the experiment, the blood flow to the left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task— the gold standard task for assessing hemispheric dominance for language. Despite showing the expected RVF advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (*r*= .18) or the visual half-field task itself (r -.004). The lack of validity here likely reflects the poor reliability of the visual field task as split-half reliability was very low (reaction time split-half: *r*= .14; accuracy split-half: *r*= .20). This illustrates the difficulty of concluding associations between laterality indices when the reliability of measurement is so low.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Hunter and Brysbaert argued that this is the result of stimuli competing during bilateral presentation, enhancing the advantage to the stimulus presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation to prevent eye movements. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) used two visual half-field tasks, word naming (192 trials) and picture naming (160 trials) in which stimuli were bilaterally presented and masked at offset. The two tasks correlated strongly (*r*= .8, 95% CI .598-.956) in a sample of 26 left-handers. Encouragingly, the LIs from these visual half-field tasks also correlated positively with laterality from functional Magnetic Resonance Imaging (fMRI) on a word generation task in 10 left-handers. However, the small sample size meant that the confidence intervals around these estimates were very wide: word naming, *r*= .63 (95% CI = .001-.902), picture naming, *r*= .77 (95% CI = .272-.943).

@@

Simultaneous presentation of different auditory stimuli to left and right ears for immediate recall was a method originally used in studies of attention, but it became evident that there was usually an advantage for verbal stimuli presented to the right ear (Kimura, 1967). Both ears project to both hemispheres, but the weaker ipsilateral pathways appear to be suppressed when the two sides compete, and so the right ear, with a contralateral projection to the left hemisphere, will have an advantage in tasks that involve language processing.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. Their recommendations were to use: (1) stop consonant-vowel (CV) syllables as stimulus materials; (2) CV stimulus pairs from the same voicing category; (3) alternating trials of voiced and unvoiced stimulus pairs; (4) free-recall instruction; (5) one stimulus pair per trial; (6) a paradigm length of 120 dichotic trials; (7) all stimulus pairs are presented to the two sides in identical frequency; and (8) to include binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded test-retest reliability of *r*= .93 for verbal responses and *r*= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

@@

In studies of face processing, it is possible to use a modification of the visual half-field paradigm in which single stimuli are viewed, composed of vertically split chimeric faces, where each half displays a different emotion or identity. Our focus here is on emotion detection in chimeric faces, where bias is usually found towards reporting the emotion shown in the left hemiface.

The reliability of the chimeric faces tasks seems promising. Early work by Levy et al. (1983) using two vertically presented chimeric faces, found split-half reliability of *r*= .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data but in a task of recognising facial identity rather than emotion. In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognise centrally presented chimeric faces, which showed different identities in the right and left visual fields. It is encouraging to note that the behavioural measure significantly correlated with the fMRI laterality index (LI) calculated from on activation in the fusiform face area (*r*= .49). However, the sample size used here was small: with 17 participants, the 95% confidence interval around the observed correlation of .49 ranged from .012 to .786.

@@

The studies reviewed so far have indicated consistent population biases in lateralisation on behavioural tasks, but also there are also substantial individual differences. One factor that has been associated with the degree of lateralisation is handedness (for a comprehensive review, see Carey & Johnstone, 2014). Neuroimaging data indicates that for language tasks these differences in lateralisation primarily reflect the contribution of a subset of strong left-handers who are right hemisphere dominant for language, rather than indicating that this relationship operates on a continuum (Mazoyer et al., 2014). This association can be used to validate behavioural measures of lateralisation: we would expect to see a departure from typical lateralisation in strong left-handers.

With regards to visual half-field tasks for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Weekes, Capetillo-Cunliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well-powered study involving 100 participants, conducted by Willemin et al. (2016), reported no hint of any effect of handedness, when coded as left and right, or when divided into left (N = 19), mixed (N = 24) and right-handed (N = 57) groups, on the visual field advantage during a lexical decision task. This result illustrates the recurring challenge of interpreting null results in this field: do they reflect a true lack of relationship, poor test reliability, inadequate power to detect small effects, or specific methodological factors? We may note that Willemin et al. asked participants to respond via manual response, whereas other studies have involved an active speech production component, which could influence task lateralisation.

@@

For chimeric faces tasks, which typically show a left visual field advantage, there again appears to be a greater departure from the typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Levy et al., 1983; Rozkowski & Snelbecker, 1982). This was also studied by Karlsson et al. (2019). When viewing a single face with a different emotion shown in either hemiface, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task that involved two chimeric faces arranged vertically, a similar pattern emerged. 73% of right-handers and 57% of non-right-handers showed a leftwards bias.

From the above-reviewed studies, it appears that handedness is related to laterality across a range of tasks (for further discussion see Carey & Johnstone, 2014). This relationship points towards a general laterality factor which influences lateralisation across tasks and behaviours, as predicted by several theoretical accounts (Behrmann & Plaut, 2015; Hellige, 1993; Kosslyn, 1987). In a study of left-handers using fMRI with five tasks, Gerrits, Verhelst, and Vingerhoets (2020) found that the just under half the participants showed either standard laterality (left-lateralised for language and praxis, right-lateralised for face memory, line-bisection and vocal emotion) or a complete reversal of this pattern, providing clear evidence against the idea that these functions are lateralised independently. It is also clear that this association is far from perfect, (Badzakova-Trajkov, Haberling, Roberts, & Corballis, 2010) and some studies have found no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). Our ultimate goal is to unpick whether dissociations between laterality indices are evidence of multiple independent lateral biases across tasks, or whether there is a more prosaic explanation in terms of poor reliability or validity of tasks.

@@

The principal goal of the current study was to evaluate the reliability and validity of an online behavioural laterality battery. Specifically, we examined whether it is possible to observe the usual laterality biases at the population level in tests of language and emotion processing and whether it is possible to achieve test-retest reliability of *r*= .65 or above using online behaviour research methods. In addition, for tasks where satisfactory reliability was achieved, we considered how many trials were needed to achieve this. The tasks included in the battery were a novel rhyme decision visual half-field task, a consonant-vowel dichotic listening task, a chimeric faces visual half-field task, and a novel finger tapping task.<!--- ZW: Would it be worth having a section on measurement of manual preference?; AP: As this is exploratory I don't want to add too much about this---> These behavioural tasks were completed twice on two different occasions.

Our first set of predictions (hypothesis 1) concerned the validity of the tasks as indicators of cerebral lateralisation. Based on previous work, we predicted that at the population level: (1) the laterality indices for rhyme decision would show a RVF advantage (indicating left hemisphere dominance for language processing); (2) the laterality indices for the dichotic listening task would show a right ear advantage (indicating left hemisphere dominance for language processing); and (3) the laterality indices for chimeric faces would show a LVF advantage (indicating right hemisphere dominance for face or emotion processing). Although exploratory, the finger tapping task was designed to provide a quantitative measure of the degree of handedness, and was expected to show a right-hand advantage in the overall sample.

Our second prediction (hypothesis 2) related to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotients of -90 or less on the Edinburgh Handedness Inventory; Oldfield, 1971) would be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population-level predictions) than right-handers. This prediction followed from observations by Mazoyer et al. (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded, “concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

Our final prediction (hypothesis 3) related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult to gather large amounts of data to establish test-retest reliability. For instance, van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of the individual measures. In the preregistration for our current study, we predicted that for right-handers, a model where the two language tasks are correlated with each other but not with the chimeric faces (model A) would be the best fit to the data, compared to either a model in which laterality indices are correlated between the language and chimeric face tasks (model B), or a base model where each task is independently lateralised (model C). Given our previous findings using fTCD to study dissociations in language laterality (Woodhead, Bradshaw, Wilson, Thompson, & Bishop, 2019), we further predicted that model fit may differ for right- and left-handers, with more dissociation between tasks in the latter group.

@@

This project was pre-registered on the Open Science Framework (OSF) before the start of data collection. The registration form along with all task materials, analysis scripts and anonymised data can be found on the OSF at https://osf.io/qsxcv/. Working versions of the dichotic listening, chimeric faces, and finger tapping tasks are available on Gorilla Open Materials: https://gorilla.sc/openmaterials/104636.

@@

In practice, our final sample exceeded 300, because we discarded participants from individual tasks for failing to meet prespecified criteria (see below). In total, 441 native English speakers were initially recruited through Prolific Academic (www.prolific.co). Participants had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss or impairment.

This study was approved by the University of Oxford's Medical Sciences Inter-Divisional Research Ethics Committee (reference number: R66638/RE002) and conducted following the principles of the Declaration of Helsinki. Participants gave informed consent and received compensation for their participation.

@@

The battery included four measures of background information and four experimental assessments of behavioural  laterality that were built and administered using the Gorilla Experiment Builder (www.gorilla.sc), a cloud-based tool for collecting data in the behavioural sciences (Anwyl-Irvine, Massonnié, Flitton, Kirkham, & Evershed, 2020). Each task is described below.

@@

The basic demographics questionnaire asked participants to report age, gender, years in education, and whether they were bilingual. We also asked questions about handedness and footedness. For handedness, we asked: which is your dominant hand (i.e. which hand do you prefer to use for tasks such as writing, cutting, and catching a ball)? For footedness, we asked: which foot do you normally use to step up on a ladder/step? For both questions, participants could answer "left", "right", or "no preference / Don't know".

@@

The Edinburgh Handedness Inventory (EHI; Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The EHI was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference. For simple contrast between left- and right-handers, we treated indices greater than 0 as right-handed, and indices less than 0 as left-handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.

@@

The rhyme decision task was developed for this project as a novel visual half-field technique designed to engage lateralised language processes. On each trial, participants were asked to decide whether a foveally presented letter string rhymed with the name of an image shown in either the LVF or RVF. With online presentation, a particular concern is a potential for eye movements away from fixation. We relied on the fact that participants had to focus initially on the central written word to do the task, with pictures subsequently displayed with brief bilateral presentation, which has been shown to give adequate control for eye movements in previous visual half-field experiments with verbal naming (Beaumont, 1982). Studies that directly monitored eye movements in this setting reported failures of fixation on only 0.5% of trials (Geffen, Bradshaw & Nettleton, 1972; but see also Bourne, 2006). Pictures of familiar objects were presented for rhyme matching to the written word, to ensure that the participant would have to covertly name the picture, thereby engaging left hemisphere speech processing.

@@

![Figure 2. An example stimulus from the chimeric faces task. In the example, “happiness” is shown on the left and “sadness” on the right. A participant who is right lateralised for emotional processing would be more likely to perceive the emotion in the left hemifield and would, therefore, respond “happiness”.](chim_example.png)

@@

autofit(ft)
```
*Note*. N is the count of participants. % is the percentage of participants.

@@

\newpage
#### Supplemental material
